---
title: "worksheet answers"
format: html
---

# Example for Women's data
# First, download the data and turn decimals into percentages
```{r}
library(tidyverse)
library(ggplot2)

WLAX <- read_csv("jtyrrell/lacrosse_women_ncaa_div1_2022_2023.csv") |> mutate(draw_pctg = draw_pctg * 100, 
                             win_pctg = win_pctg * 100)
```

# Next, plot the regression model 
```{r}
ggplot(WLAX, aes(x = draw_pctg, y = win_pctg)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Winning the Draw vs. Winning the Game",
       x = "Draw Win Percentage",
       y = "Game Win Percentage") +
  theme_minimal()
```

# Next, calculate the regression line
```{r}
model <- lm(win_pctg ~ draw_pctg, data = WLAX)
summary(model)
```

# Interpret these results (what do these numbers mean?)
# a.) draw percentage coefficient (2.4478): 
for every 1% increase in draw win percentage, game win percentage increases by 2.4478 %
# b.) p value (p = 0.0000000000000002)
the low P value depicts that draw percentage significantly affects win percentage
# c.) R^2 value (R^2= 0.5016)
the model explains 50.16% of the variation in win percentage
# d.) Residual Standard Error (RSE = 14.95)
on average, the model's predictions are off by about 14.95 units (percentage points)

# Now you try with the mens data!
# men's regression model
```{r}
MLAX <- read_csv("jtyrrell/lacrosse_men_ncaa_div1_2022_2023.csv") |> mutate(faceoff_win_pct = faceoff_win_pct * 100,
                                                                          win_loss_pctg = win_loss_pctg * 100)

ggplot(MLAX, aes(x = faceoff_win_pct, y = win_loss_pctg)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Winning the Draw vs. Winning the Game",
       x = "Draw Win Percentage",
       y = "Game Win Percentage") +
  theme_minimal()
```
# 1.) Calculate the men's regression line:
```{r}
model2 <- lm(win_loss_pctg ~ faceoff_win_pct, data = MLAX)
summary(model2)
```

# 2.) Interpret the results (what do these numbers mean?)
# a.) draw percentage coefficient
for every 1% increase in draw win percentage, game win percentage increases by 1.5024 %

# b.) p value
p = 0.000001637, depicting that draw percentage significantly affects win percentage

# c.) R^2 value
R^2 = 0.2815, so the model explains 28.15% of the variation in win percentage

# d.) Residual Standard Error
RSE of 17.49 means that, on average, the model's predictions are off by about 17.49 units (percentage points)

# We don't have to analyze men and women separately, we can do them together! Continue with this data set that combines both the men's and women's data
```{r}
WMLAX <- read_csv("jtyrrell/lacrosse_men_and_women_ncaa_div1_2022_2023.csv") |> mutate(faceoff_win_pct = faceoff_win_pct * 100,                                                              win_loss_pctg = win_loss_pctg * 100)

WMLAX$Division <- as.factor(WMLAX$Division)

ggplot(WMLAX, aes(x = faceoff_win_pct, y = win_loss_pctg, color = Division)) +
  geom_point(alpha = 0.7) +  
  geom_smooth(method = "lm", formula = y ~ poly(x, 1), se = TRUE) +  
  labs(title = "Regression Model: Draw Wins vs. Game Wins",
       x = "Draw Win Percentage",
       y = "Game Win Percentage") +
  theme_minimal()
```

# 1.) Calculate the interaction model
```{r}
WMmodel <- lm(win_loss_pctg ~ faceoff_win_pct * Division, data = WMLAX)
summary(WMmodel)
```


# 2.) Tnterpret results (what do these numbers mean?)
# a.) How do the regression lines differ for men's and women's lacrosse?
The women's slope is steeper, depicting that draw win percentage has a stronger impact on game with percentage compared to the men.
The interaction term (β = 0.9454, p = 0.0084) shows that the effect of winning faceoffs/draws is stronger for women’s teams than for men’s teams. 
For every 1% increase in draw win percentage, women’s win percentage increases by (1.5024 + 0.9454 = 2.4478%), while men’s win percentage increases by only 1.5024%.

# b.) Which division (men's or women's) seems to show a stronger relationship between draw wins and game wins?
The women division shows a stronger relationship due to their steeper slope and higher R^2 value

# c.) What does this tell you about the importance of winning faceoffs/draws in lacrosse games?
Overall, winning the draw/faceoff in the women division is more crucial and important to the games overall outcome than in the men division