---
title: "Boston Marathons - ANOVA"
format: html
---

The Boston Marathon is an annual marathon traditionally held on Patriots' Day, the third Monday of April. Begun in 1897, the event is one of the world's oldest annual marathons and ranks as one of the world's most prestigious road racing events. The Boston Marathon attracts runners from all over the world, featuring thousands of participants and extensive spectator support, embodying the spirit of community and endurance.

We will be investigating the results from the 2023 Boston Marathon - which consists of 26,598 runners that completed the race.  In addition to the `name`, `gender`, and `age_group` of the runner, the `finish_net_minutes` variable shows the race result time of the runner that is timed from when they cross the starting gate at the beginning of the race to the finish line. The full data (boston_marathon_2023.csv) is available at [SCORE Data Repo](https://data.scorenetwork.org/running/boston_marathon_2023.html){target="_blank"} and contains many more variables associated with the runner information and times.

0. Load the following packages: `tidyverse`, `multcomp`

```{r}
library(tidyverse) |> suppressPackageStartupMessages()
library(multcomp)|> suppressPackageStartupMessages()
```


1. Load the data into R. Store it as an object names `bos_marathon`.

```{r}
bos_marathon <- read.csv("https://data.scorenetwork.org/data/boston_marathon_2023.csv") |>
  mutate(age_group = str_replace_all(age_group, "-",".."))
```


2. Construct a histogram of the distribution of finish times for all participants. If using `ggplots`, use `bins=200` in the options. If using base R's `hist` function, use `breaks=200`. (Note that setting the bins/breaks to be a large number is only feasible because the sample size is so large.)

```{r}
# use

ggplot(bos_marathon, aes(x = finish_net_minutes)) +
  geom_histogram(bins = 200)

# or

hist(bos_marathon$finish_net_minutes, breaks = 200)
```


3. Notice that, while predominately right skewed with a major mode around 220 minutes, there seems to be a clear spike at approximately 175 minutes. Comment on some reasons why that might occur.

_Answers will vary - but the main reason this seems to happen is due to displaying both Men and Women participants. Essentially we have a mixture of the two distributions displayed here._

4. boxplot min ~ gender
5. Fit finish_net_minutes ~ gender
6. Calculate and interpret R-Squared for this model
7. boxplot min ~ age
8. model min ~ age
9. R^2
10. Interactions?
 - interaction plots
 - model
11. Final model
 - significance testing
 - Tukey HSD


```{r}
bmar_2way_int_mod <- aov(finish_net_minutes ~ gender + age_group + gender:age_group, data = bos_marathon)
```

```{r}
library(multcompView)
Letters <- multcompLetters4(bmar_2way_int_mod, TukeyHSD(aov(bmar_2way_int_mod)), 
reversed = TRUE)

```


```{r}
marathon_mns <-   
  bos_marathon |>
    group_by(gender,age_group) |>
    summarise(mean=mean(finish_net_minutes))

  
  
tukey.cld <- as.data.frame.list(Letters$`gender:age_group`)

tukey.cld$Levels <- rownames(tukey.cld) ; rownames(tukey.cld) <- NULL

tukey.cld <- dplyr::select(tukey.cld,Levels,Letters)
tukey.cld <- separate(tukey.cld, col = "Levels", into = 
c("gender","age_group"), sep = ":",remove = T)
tukey.cld
```



