---
title: "Cleaning Module Key"
format: html
---

1. Load in necessary packages
```{r}
library(tidyverse)
```

2. Read in Tremblant_Data.csv
```{r}
library(readr)
Tremblant_df <- read_csv("Tremblant_Data.csv")
```

3. Write a function that converts time to seconds
```{r}
fix_time <- function(time){
  n = length(time)
  Diff = time[2:n]
  t1 = time[1] |> ms() |> period_to_seconds()
  Diff <- parse_number(Diff)
  Diff <- t1 + Diff
  return(c(t1, Diff))
}
## this function works for times over a minute
```

4. Fix the Total column using the function above
```{r}
Tremblant_df$TOTAL <- fix_time(Tremblant_df$TOTAL)
```

5. 5.	Separate Run 1 into two columns for Run 1 time and Run 1 rank
```{r}
Tremblant_df <- Tremblant_df |>
  separate(col = `RUN 1`, into = c("Run_1_Time", "Run_1_Rank"),
           sep = "\\(")
```

6. Do the same as above for Run 2
```{r}
Tremblant_df <- Tremblant_df |>
  separate(col = `RUN 2`, into = c("Run_2_Time", "Run_2_Rank"),
           sep = "\\(")
```

7. 7.	Use a function to make sure there are only numeric values in the Pr, Run 1 Rank, and Run 2 Rank columns
```{r}
Tremblant_df <- Tremblant_df |>
  mutate(Run_1_Rank = parse_number(Run_1_Rank)) |>
  mutate(Run_2_Rank = parse_number(Run_2_Rank))
```

8. 8.	Use your function to fix time on Run 1 time and Run 2 time
```{r}
Tremblant_df$Run_1_Time <- fix_time(Tremblant_df$Run_1_Time)
Tremblant_df$Run_2_Time <- fix_time(Tremblant_df$Run_2_Time)
```

9. Rename variables logically. Answers will vary.
```{r}
Tremblant_df <- Tremblant_df |> rename(Final_Rank = RANK,
                                  Run2_Order = N,
                                  Run1_Order = BIB,
                                  Total_Time = TOTAL,
                                  Run1_Time = Run_1_Time,
                                  Run1_Rank = Run_1_Rank,
                                  Run2_Time = Run_2_Time,
                                  Run2_Rank = Run_2_Rank)
```

10.	Order the variables logically. Answers will vary.
```{r}
Tremblant_df <- Tremblant_df |>
  select(4, 5, 3, 8, 9, 2, 10, 11, 6, 1, 7)
```

