---
title: "FigureSkatingAnova_Solutions"
Authors: Faith Rhinehart
execute: 
  echo: true
  warning: false
  error: true
format: 
  html:
    embed-resources: true
---

Figure skating is a sport where people perform routines on ice to music, combining athletic skill with artistic expression. Skaters compete in different categories like men's singles, women's singles, pairs (a man and a woman skating together), and ice dance (which focuses more on rhythm and movement). Each skater or team performs two routines: the short program and the free skate. The short program is a shorter routine with a list of required moves that every skater must include. Judges look at how well each move is done and how smoothly everything fits with the music. The free skate is longer and gives skaters more freedom to show their personal style and creativity. They still need to include certain types of moves, but they can choose how to put them together. Judges give scores for both routines based on how difficult the moves are, how well they’re done, and how artistic the performance is. The scores from both routines are added together, and the skater or team with the highest total wins.

Read in the dataset:
```{r}
library(ggplot2)
library(tidyverse)
library(here)
library(dplyr)
menSkating <- read_csv(here::here("frhinehart/Olympic Figure Skating/Module/MensFigureSkating.csv"))
```

1) Create a box plot graph with `Year` on the x-axis and `TP` (Total Points) on the y-axis. Group by Year to get a box plot for each year. Describe what you see. 
```{r}
ggplot(data = menSkating, aes(x = Year, y = TP, group = Year)) +
  geom_boxplot()
```

The means of each year's box plots are different. 

We would like to determine if these means have a significant difference. To do this we will use ANOVA (a test that analyzes the differences among group means) But first we will need to check for assumptions.  

2)

The 3 assumptions are Independence (if the samples are collected independently from each other), Normality (if the data is normally distributed; not skewed), and Constant Variance (the difference between observed and predicted values is equal among data points). These assumptions allow us to obtain reliable data.

Using the box plots, comment on the 3 assumptions. You can assume the data was collected independently
Independence - Yes
Normality - Yes, the box plots are not significantly skewed
Constant Variance - Yes the whiskers are roughly the same.

Now that we have checked the assumptions we can preform ANOVA. 

3) 

a - State the null hypothesis and alternative hypothesis

Null - There is no significant difference in the mean total points throughout the years 

Alternative - There is a least one significant difference in the mean total point throughout the years

b - Preform an Anova to determine if there is an significant difference of `TP` (Total Points) between the years using the aov() function. You will need to change `Year` to a factor. 
```{r}
anova_result <- aov(TP ~ as.factor(Year), data = menSkating)
summary(anova_result)
```

c - Report the F-Staistic, degrees of freedom, and p-value

F-Statistic = 20.95
P-value = 5.39e-13
df = 4 and 115

d - Use the p-value to make a conclusion

reject the null hypothesis

There is significant evidence to conclude (F-Statistic = 20.95, P-value = 5.39e-13, df = 4 and 115) that there is a least one significant difference in mean total point between the years

In order to find out what years have a significant difference we can preform a Post Hoc Tukey Test. 

4) 

a - Preform a Post Hoc Tukey test using the TukeyHSD() function to determine between which years there is a significant difference in scoring. Report those years. Report the years that are not significant.
```{r}
tukey_result <- TukeyHSD(anova_result)
tukey_result
```

Significant:
2014 and 2006
2018 and 2006
2022 and 2006
2018 and 2010
2022 and 2010
2018 and 2014
2022 and 2014

Not significant:
2010 and 2006
2014 and 2010
2022 and 2018

The Tukey Test also performs a confidence interval. If the confidence interval does not include 0 then we know that it supports the p-value.

b - Interpret a confidence interval for 2022 and 2014 comparison

With 95% confidence, the true difference in mean total points between 2022 and 2014 lies between 61.50194 and 0.0005747. Since this interval does not include zero, it further supports the conclusion that there is a significant difference between the groups.

Optional:

A Pair Wise Comparison is used for significance testing. Each column is assigned a letter to indicate if it is significantly different from another column (for this case each year will represent a column). If two columns don't have a matching letter then they are significantly different. We can use the TukeyHSD output to do this. 

We arrange the columns (whatever variable is on  th x-axis of your plots) typically from largest sample size to smallest sample size but since the years all have the same sample size we will put them in order from largest year to smallest year). This will be our letter table. 

2022 --- a
2018 --- a
2014 ---  b 
2010 ---  b c 
2006 ---    c 

To make the table, we will compare the significance between each of the years with each other. If it is not significant, both years gets a letter (following abcd... order). 

For example, starting with 2022...
Any year that is not significant with 2022 gets the letter "a". Place a letter "a" at 2022 and 2018 since they are not significant with each other. No other year will have an "a" because they are significant with 2022. 

Then we move to 2018 with the letter "b". Since all years have significance with 2018 none of them get a letter "b", we move onto 2014 using "b".

a - Place a "b" with any year that does not have significance with 2014 (remember to also place one at 2014). If there is none move onto the next year with "b"

b - Place a "c" with any year that does not have significance with 2010 (remember to also place one at 2010). If there is none move onto the next year with "c"
Note: if you already paired two years together as not significant you don't pair them again with a new letter. For example you already paired 2014 and 2010 with "b", you do not pair them again with "c" (if you do it can lead to confusion that 2006 and are not significant when they are.)

c - Place a "d" with any year that does not have significance with 2006 (remember to also place one at 2006).

d - Did you get the same results as what you got in 6a? 

Yes

