---
title: "NBA Wingspan & Performance Solutions (no webscraping)"
format: html
---

In professional basketball, physical traits can have a major impact on how a player performs. One key trait that often draws attention from scouts and analysts is **wingspan**, the distance from fingertip to fingertip with arms fully extended.

This worksheet explores the question:
How does wingspan, especially wingspan relative to height, relate to player performance in the NBA?

You’ll analyze data combining NBA player profiles (including height and wingspan) with in-game statistics from the 2024–25 season. A key variable is **wingspan_advantage**: the difference between a player’s wingspan and height, often viewed as a potential edge in defense and rebounding, but a potential disadvantage in shooting.

Throughout this worksheet, you’ll clean and combine messy datasets, visualize relationships, and explore meaningful patterns, just like a data analyst working for an NBA front office. Your goal is to investigate whether a player’s wingspan advantage is linked to any performance metrics, and to think critically about how physical traits might (or might not) translate to on-court impact.



# 0. Load the Following Packages

```{r}
#| message: false
library(tidyverse)
library(rvest)
library(janitor)
```

# 1. Load in CSV file

Load in the `nba_wingspan_2025.csv` from the data folder.

```{r}
wingspan <- read_csv(here::here("colney/nba_wingspans/data_repo_submission/nba_wingspan_2025.csv"))
```

# 2. Clean the Dataset

Clean the dataset using the following packages from tidyverse: `dplyr`, `tidyr`, `readr` and `stringr`.

Once you are finished cleaning this dataset, these should be the following variables.

| Variable | Type | Description |
|------------------------|------------------------|------------------------|
| `name` | chr | Full name of the NBA player |
| `team` | chr | Three-letter abbreviation of the team the player is on |
| `position` | chr | Player's primary on-court position in abbreviated form |
| `height_inches` | num | Player's height in inches |
| `wingspan_inches` | num | Player's wingspan in inches |
| `wingspan_advantage` | num | Difference between wingspan and height in inches (wingspan - height) |

a.  Convert the `height` variable (which is currently a character string like 6'4") into a new numeric variable called `height_inches` that represents each player's height in total inches. **Hint** using helper variables (such as separate variables for feet and inches) can make this process easier.

```{r}
wingspan <-
  wingspan %>%
  separate_wider_delim(
    cols = height,
    names = c("feet", "inches"),
    delim = "'"
  ) %>%
  mutate(
    feet = parse_number(feet),
    inches = parse_number(inches),
    height_inches = feet * 12 + inches
  ) %>%
  select(-feet, -inches)
```

b.  Do the same thing as above, but with the `wingspan` variable. Create a new numeric variable called `wingspan_inches` that represents each player’s wingspan in total inches.

```{r}
wingspan <-
  wingspan %>%
  separate_wider_delim(
    cols = wingspan,
    names = c("feet", "inches"),
    delim = "'"
  ) %>%
  mutate(
    feet = parse_number(feet),
    inches = parse_number(inches),
    wingspan_inches = feet * 12 + inches
  ) %>%
  select(-feet, -inches) %>%
  relocate(wingspan_advantage, .after = wingspan_inches)
```

c.  Extract the player's position (e.g. "SG" or "C") from the `name` variable and store it in a new variable called `position`, so that the position is no longer part of the name column.

```{r}
wingspan <-
  wingspan %>%
  separate_wider_delim(
    cols = name,
    names = c("name", "position"),
    delim = " | "
  )
```

d.  Remove players who are not currently on a team. In the `name` column, these players do not have a three-letter team abbreviation at the end of their name (e.g. "LeBron JamesLAL" vs. "LeBron James"). Then, split the name and team abbreviation into two separate variables: `name` and `team`. **Hint:** Use stringr functions to complete this step.

```{r}
wingspan <-
  wingspan %>%
  filter(
    str_detect(name, pattern = "[A-Z]{3}$")
  ) %>%
  mutate(
    team = str_extract(name, pattern = "[A-Z]{3}$"),
    name = str_remove(name, pattern = "[A-Z]{3}$")
  ) %>%
  relocate(team, .after = name)
```

# 3. Load Data From Basketball Reference

**Why Per 100 Possessions:** Per 100 possessions stats are often preferred over per game stats in basketball because some teams have more possessions simply due to pace. These stats adjusts for that by standardizing performance across the same number of plays, making it easier to compare players fairly and evaluate efficiency and impact.

a.  Load the data from the `nba_per100possessions_2025.csv` file from the data folder.

```{r}
per_100_poss <- read_csv(here::here("colney/nba_wingspans/data_repo_submission/nba_per100possessions_2025.csv"))
```

b.  To tidy the dataset by completing the following steps:

**First**, remove duplicate player entries by keeping only the row that represents a player’s full season total. (For example, Luka Dončić was traded mid-season, so he appears three times: once for his stats with DAL, once with LAL, and once for his total 2024–25 season stats.) **Hint:** Use `group_by()` and `ungroup()` to help identify duplicates. Also, games played will be the highest for rows that represent a full season total.

**Optional:** Use `select()` to keep only the variables you're interested in.

```{r}
per_100_poss <-
  per_100_poss %>%
  group_by(player) %>%
  slice_max(order_by = g, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  # remove any unwanted rows
  slice(-361)
```

c.  Do the same steps as in parts **a** and **b**, but this time use the `nba_shooting_2025.csv` file from the data folder.

```{r}
shooting <- read_csv(here::here("colney/nba_wingspans/data_repo_submission/nba_shooting_2025.csv"))
```

```{r}
shooting <-
  shooting %>%
  group_by(player) %>%
  slice_max(order_by = g, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  # remove any unwanted rows and cols
  slice(-361) %>%
  select(-2)
```

d.  Combine the two cleaned datasets from Basketball Reference into a single dataset.

```{r}
bball_ref <-
  per_100_poss %>%
  left_join(shooting, by = "player")
```

# 4. Combine Datasets

Combine the cleaned wingspan dataset from part 2, with the combined Basketball Reference dataset you created in part 3.d.

In the 'nba_wingspan_2025.csv' file, two names were intentionally misspelled compared to their spellings on Basketball Reference.

-   CSV spelling → Correct spelling (Basketball Reference)
-   Jimmy Butler III → should be Jimmy Butler
-   Alperen Sengun → should be Alperen Şengün

Keep this in mind as you work through the questions.

Combine the datasets using **two** different strategies.

a.  Use `left_join()` to combine the datasets. Do you notice the misspelled names, what happens as a result of the misspelled names?

```{r}
wingspan %>%
  left_join(bball_ref, by = c("name" = "player"))
```

b.  Use `stringdist_join()` from the `fuzzyjoin` package to combine the datasets. How does the result differ from using `left_join()`?

```{r}
stringdist_join(
  wingspan, bball_ref,
  by = c("name" = "player"),
  method = "osa",
  max_dist = 2,
  mode = "left"
) %>%
  select(-player) %>%
  
  
  # run this to show players that it doubles up
  count(name) %>%
  filter(n > 1)
```

c.  After comparing the differences, use `mutate()` to correct the two misspelled names in the wingspan dataset. Then, recombine the datasets and this time there should be no mismatches.

```{r}
combined <- 
  wingspan %>%
  mutate(
    name = if_else(name == "Alperen Sengun", "Alperen Şengün", name),
    name = if_else(name == "Jimmy Butler III", "Jimmy Butler", name)
  ) %>%
  left_join(bball_ref, by = c("name" = "player"))
```

# 5. Explore the Combined Dataset

Use the newly created dataset to investigate potential relationships between a player's physical traits and their performance on the court.


a. Create a histogram of the `wingspan_advantage` variable to see how common different levels of advantage (or disadvantage) are across players in the NBA.
```{r}
combined %>%
  ggplot(., aes(x = wingspan_advantage)) +
  geom_histogram(binwidth = 1)
```

b. Create a scatterplot using `wingspan_advantage` as the explanatory variable and `blk` as the response variable. What kind of relationship, if any, do you observe? Are there any outliers?
```{r}
combined %>%
  ggplot(., aes(x = wingspan_advantage, y = blk)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE)
```

c. Create a scatterplot using `wingspan_inches` as the explanatory variable and `3pt_percentage` as the response variable. Include a regression line, a title, and labels for the x and y axes. Then separate the plot by `position`. What patterns or relationships stand out within or across positions?
```{r}
combined %>%
  ggplot(., aes(x = wingspan_advantage, y = `3pt_rate`)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~position)
```

d. **Optional:** Investigate any other combination of physical traits and on-court performance metric you find interesting. Create a visualization to explore the relationship and try to explain any patterns or outliers you observe.
```{r}

```









