---
title: "Difference of Means CI"
output: html_document
date: "2023-07-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggiraphExtra)

boston <- read_csv("data/boston_marathon_2023.csv")

# checking the sample sizes for each age group
boston%>%
  group_by(age_group, gender)%>%
  summarise(number = n())

# filtering and subsetting
boston<- boston%>%
  filter(age_group == "18-39" | age_group == "40-44" | age_group == "45-49")


# Possibly: ADD ANOTHER AGE GROUP AND MAKE APPROPRIATE CHANGES TO QUESTIONS ?
```

# Module 

## ANOVA for Men with 3 age groups 
```{r}
bostonMEN <- boston%>%
  filter(gender == "M")
menAOV <- aov(finish_net_minutes ~ age_group, bostonMEN)
summary(menAOV)
model.tables(menAOV, type = "means")
model.tables(menAOV, type = "effects")
```

## What are the mean times (in minutes) of each age group?

## Which age group is most abnormal in terms of effect?

```{r}
TukeyHSD(menAOV)
```
### Is the difference larger between age groups 18-39 and 40-44 or age groups 40-44 and 45-49?

### Interpret a 95% confidence interval for the larger difference of means 

-------------------------------------------------------------------------------------------------
## ANOVA for Women with 3 age groups 
```{r}
bostonWOMEN <- boston%>%
  filter(gender == "M")
womenAOV <- aov(finish_net_minutes ~ age_group, bostonWOMEN)
summary(womenAOV)
model.tables(womenAOV, type = "means")
model.tables(womenAOV, type = "effects")
```

## What are the mean times (in minutes) of each age group?

## Which age group is most abnormal in terms of effect?


```{r}
TukeyHSD(womenAOV)
```
### Is the difference larger between age groups 18-39 and 40-44 or age groups 40-44 and 45-49?

### Interpret a 95% confidence interval for the larger difference of means.

### Is the larger difference of means among age groups the same for men and women? Support your answer with reasoning.
